---
title: "NEI_its_Reliability"
author: "Yansen Chen"
date: '2022-11-11'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Base information

This page shows how to use R code for calculating the nitrogen efficiency index (NEI) and its reliability with example data.

The NEI combining the (G)EBV of N intake (NINT, Kg/d), milk true protein N (MTPN, Kg/d), and milk urea N yield (MUNY, g/d) through the selection index. PS: the units of the 3 features are different 

The features (NINT, MTPN, MUNY) were divided into 2 classes: primiparous and multiparous (lactations 2 to 5) and shown as NINT1, MTPN1, MUNY1, NINT2+, MTPN2+, and MUNY2+ (6 traits in total)

The index weights (a) of the 6 studied traits were calculated by selection responses, which assumed that the genetic selection responses for NINT, MTPN, and MUNY were 0, 1, and -1, respectively. This assumption allows genetic selection for NEI to keep NINT unchanged, increase MTPN, and decrease MUNY. 



## 2. Read example data
Please download the example data (2 files) from https://github.com/Yansen0515/GP_NEI.
Then put them in your work directory
```{r}
# install.packages("data.table")
library(data.table)# please download it before using it
db <- fread("example_data.csv") # read example data
head(db) # check the example data
str(db)
G <- as.matrix(fread("A_G.csv")) # read genetic (co)variance
G
```

#### This example data includes two martixs: db and G
#### The db martix (1,000 x 13) includes 1000 animals with EBV  of 6 traits and their reliabilities.
#### The G martix (6 x 6) includes genetic (co)variance component of 6 traits

## 3. The NEI

Calculating the NEI base on the EBV of 6 traits, the index weights (a) first need to get. 

### 3.1 Index weights (a)
```{r}
## Create the transformation matrix (T), which is used to transform six traits into three features 
## Half of each feature in primiparous and multiparous (lactations 2 to 5)
T = matrix(c(0.5, 0,    0,   0.5,  0,    0,
             0,   0.5,  0,   0,    0.5,  0,
             0,   0,    0.5, 0,    0,   0.5), ncol = 6, byrow = T)

d_g <- diag(diag(G)) # get the variance for six traits

#  Standardized genetic (co)variance components because the units of the 6 traits are different
Gs <- solve(sqrt(d_g)) %*% G %*% solve(sqrt(d_g)) 

Gs2 = T %*% Gs %*% t(T) # Get the transformed Gs2 matrix

Ds2 = diag(diag(Gs2))
T2 = solve(sqrt(Ds2)) %*% T # Standardized transformation matrix T

Gs3 = T2 %*% Gs %*% t(T2)   # Get the standardized transformation matrix Gs3

r = matrix(c(0,1,-1), ncol = 1) # create the selection response vector (r) of 3 features

a = solve(Gs3) %*% r # get the standardized index weight of 3 features
a2 = t(a)%*%T2       # Get the standardized index weight of 6 traits
a_orig = t(a2 %*% solve(sqrt(d_g))) # Get the index weight of 6 traits
a_orig

```

### 3.2 NEI values

```{r}
db$NEI = as.matrix(db[,2:7]) %*% a_orig
head(db)
```
## 4. Reliability of NEI


```{r}
G <- as.matrix.data.frame(G)
Vg = t(a_orig) %*% G %*% a_orig # Maximum  genetic variance
sREL_six <- sqrt(db[,8:13]) # square root of relibility (REL) for each trait

# using the loop to get the relibility (REL) of NEI
REL_NEI <- c()
for (i in 1:nrow(sREL_six)) {
  w_tem <- a_orig * t(sREL_six[i])
  Vanimal = t(w_tem)%*%G%*%w_tem
  REL_tem = Vanimal/Vg
  REL_NEI = rbind(REL_NEI, REL_tem)
  
}

df <- cbind(db, REL_NEI) #put all results together
colnames(df)[15] <- "Relibility_NEI"
head(df)

```


## Finally, we have NEI and its reliability


For more theoretical and formula details, please read the following article?


Chen, Y., H. Atashi, C. Grelet, S. Vanderick, H. Hu, and N. Gengler. 2022. Defining a nitrogen efficiency index in Holstein cows and assessing its potential impact on the breeding program of bulls. J. Dairy Sci. 105: 7575-7587. https://doi.org/10.3168/jds.2021-21681.
